ptr-tidy: Automatic refactoring of raw pointers in C++

### 2020-12-12 proposal.tex
add risk assessment & more background, fix review comments

    \title{ptr-tidy: Automatic refactoring of raw pointers in C++}
    \author{Artem Usov}
    \date{18 December 2020}
    \maketitle
    \tableofcontents
    \newpage

@@

    Memory safety errors occur when languages are used that place the tasks of memory management with the programmer, such as C and C++.
    However simply not using these languages is not an option since their inherent danger makes them highly performant, and therefore the only option for a project such as a web browser.

@@

    Engineers at both Microsoft~\cite{Thomas2019} and Mozilla~\cite{Hostfelt2019} converge on Rust~\cite{Balasubramanian2017} as a possible solution.
    Rust is a systems language that offers the same performance as C and C++~\cite{Lin2016}, however its linear type system and memory ownership model also guarantee memory safety.
    The rewriting of a program in a new language is a colossal undertaking, especially in language that has a reputation for being difficult to learn.
    We therefore propose an alternative partial solution to this problem by using a concept from Rust.

@@

    Generally pointers are used to give the program access to a resource that cannot be directly included in the program itself, such as a file or a thread.
    However Stroustrup claims that pointers to objects allocated on the free store is dangerous and a \emph{plain old pointer}, or raw pointer as we will refer to them, should not be used to represent ownership~\cite{Stroustrup2018}.

@@

    In \autoref{lst:manual-leak} we can see that when using a raw pointer, there are two cases in which the programmer will never free the allocated memory.
    Instead we can \emph{smart pointers}~\cite{Dimov2003} and the concept of \emph{RAII (Resource Acquisition Is Initialisation)} to create resource handles which automatically eliminate resource leaks with no added overhead.
    Firstly there are \emph{unique pointers} which are a pointer that has only a single unique owner at any time.
    We can therefore fix the cases in which we leak the memory using a unique pointer which can be seen in in \autoref{lst:first-example-unique}.
    In the comments, we can see all the locations where \emph{p} exits the scope of the function and where the compiler will insert delete statements, now ensuring that the resource will never get leaked.
    The ownership of a unique pointer can be transferred between variables using \mintinline{c++}{std::move} which moves the ownership from one variable to another.

@@

    There also exists a \emph{shared pointer} for resources which may not have a single unique owner.
    Reference counting is used to ensure that after the last owner loses access to the resource, it will finally be freed.
    Shared pointers are suitable to be used in most situations, except in structures which have cyclic references \autoref{subsec:shared-pointer-cycle-analysis}.

@@

    We therefore propose to create an automatic refactoring tool \emph{(ptr-tidy)}, that statically analyses C++ code and aims to refactor usage of raw pointers into smart pointers.
    It will involve a conservative analysis that first and foremost tries to identify raw pointers that only have a single unique owner at any point in the program.
    If this cannot be guaranteed, then a more conservative refactoring to a shared pointer will be applied.

    Given that the language standards are mostly backwards-compatible~\cite{Winters2020}, we can apply these automatic refactoring without changing the runtime behaviour compared to the original programs.

@@

    \emph{"Can the translation of raw pointers into smart pointers using escape analysis be used to make existing C++ code more modern and extendable?"}

    \emph{"Can the tool identify situations in which unique pointers should be used reliably?"}

@@

    C++ as a language is massive, and the first capability our tool will need is to be able to parse C++ code.
    Doing this ourselves would be an undertaking that would likely span longer than the time span of the whole project, therefore it was decided early to use an existing parser.
    The most capable and mature C++ parsers will naturally be the parsers used by the most popular compilers.
    This leaves us therefore with a choice between GCC~\footnote{https://gcc.gnu.org/} and Clang~\footnote{https://clang.llvm.org/}.
    The literature~\cite{duffy2014} is fairly clear on the fact that the design of GCC makes it unsuitable for integration with other projects.
    Clang on the other hand allows us access to the Clang abstract syntax tree (AST) and parser internals through consumers and visitors.

    Clang is a front-end to the LLVM framework~\cite{Lattner2004} which transforms C++ code into the LLVM intermediate representation (IR).
    % TODO define SSA
    The LLVM IR is a language-independent, static single assignment (SSA) representation of the program and will be the target on which we apply our analysis to identify refactoring opportunities.
    We believe this choice will lead to several benefits in terms of the analysis power and future usage of the project:
    \begin{itemize}
        \item First, the usage of SSA allows us to efficiently generate a definition-usage graph, therefore allowing an efficient analysis of all the usages of a pointer to identify its number of owners.
        \item Secondly, by separating the parsing and analysis, we make our analysis stage reusable by any other LLVM frontends.
    \end{itemize}

@@

    \begin{listing}
        \inputminted{c++}{code/code-to-ast.cpp}
        \inputminted[fontsize=\footnotesize]{text}{code/code-to-ast.txt}
        \centerline{\includegraphics{images/code-to-ast}}
        \caption{Example of converting C++ code to Clang AST. All diagrams show equivalent representation of a small C++ code snippet}
        \label{lst:code-to-ast-example}
    \end{listing}

@@

    The escape analysis algorithm described by \citet{Choi1999} is used as the default escape analysis algorithm in the Oracle Java Virtual Machine implementation~\footnote{https://docs.oracle.com/en/java/javase/15/vm/java-hotspot-virtual-machine-performance-enhancements.html} and is widely cited so forms the basis of our investigation.
    Due to how similar C++ and Java are in terms of their Object-Oriented features, will analyse their algorithm as well as the implementations of escape analysis in the LLVM and Go compilers.

@@

    Go is another language that uses a escape analysis to allocate objects on the stack rather than the heap.
    Their escape analysis algorithms is not defined in the specification of the language, however we can look at the source~\footnote{https://github.com/golang/go/blob/master/src/cmd/compile/internal/gc/escape.go} to determine what algorithm they use.
    Fortunately the comments give a summary of their algorithm, which involves constructing a weighted graph where vertices represent variables, and edges represent assignments between variables.
    They walk the graph, looking for assignment paths that violate the invariants of a variable being allocated on the stack and mark those as requiring heap allocation
    We can see that this approach is highly similar to the use of connection graphs by \citet{Choi1999}, however there is lacking documentation on how they perform interprocedural analysis so their analysis is perhaps weaker.

    Finally, LLVM also uses escape analysis to remove any unnecessary heap allocations.
    The escape analysis algorithm is also not formally defined, however we can also look at the source code~\footnote{https://github.com/llvm/llvm-project/blob/master/llvm/lib/Analysis/CaptureTracking.cpp}.
    We see that the \mintinline{c++}{PointerMayBeCaptured} function uses the fact that the single static assignment (SSA) in the LLVM IR allows for efficient creation of definition-usage chains.
    For any defined value, such as a value initialising a variable, we can traverse all its usages, such as copying to another variable.
    To determine if an object escapes, we analyse all its usages, and all their usages and so on.
    For each usage, we then determine if it is an instruction which would cause the object to escape.
    This leads to a roughly similar analysis as the work by \citet{Choi1999} when considering the intraprocedural analysis alone, as we informally explore a graph of the nodes that represent objects and nodes that obtain references to objects by using the original value in a copy instruction.
    However, its assumption about different instructions are conservative, and it also has no interprocedural analysis, making it the weakest analysis so far.

@@

    Swift remains mostly a language for the Apple MacOS and iOS environments for application creation, however it has seen some usage for system programs by companies such as IBM~\footnote{\url{https://github.com/ibm-swift}}.
    Ada or its subset language SPARK are used extensively in certain industries to create verifiable programs for critical systems such as in avionics.
    It is perhaps most well-know for being used in the F-22 fighter aircraft~\cite{malas1997f}, however the use of Ada for those systems show its other weakness in productivity time as it was the cause for several delays in the project, and so it largely remains unused in areas which do not call for such meticulous safety.

    Rust has been the most language in attractive widespread adoption as a systems language.
    Its popularity could be quantified by considering that it is being considered for usage in writing Linux kernel modules~\footnote{\url{https://lore.kernel.org/lkml/CAKwvOdmuYc8rW_H4aQG4DsJzho=F+djd68fp7mzmBp3-wY--Uw@mail.gmail.com/T}}, which cannot be said for C++.
    Following the advice given by many that most systems programs should be written in Rust from now on, we see a potential future usage of our work in the creation of a C++ to Rust transpiler.

@@

    Garbage collected languages encompass some of the most popular programming languages such as Java and Python.
    They are considered to be easier languages than C++ since they remove the need for manual memory management by the programmer.
    In C++, if memory is improperly freed, it gets leaked and can eventually cause out of memory errors.
    Garbage collectors instead run periodically and scan the memory of the program to find objects in memory which were allocated but are no longer used anymore, i.e leaked and then free the leaked objects.
    Languages such as Java~\cite{Tauro2012} use this fully by allocating almost every object on the heap and not freeing any object, instead relying on the garbage collector to clean up.
    They are therefore considered to be easier and more productive to use.

    Garbage collection does also exist for C++ such as the Boehm-Demers-Weiser conservative garbage collector~\cite{boehm2002garbage}.
    So why is garbage collection not used in every language?
    Garbage collection is not zero-cost as the garbage collection algorithm typically stops the execution of the running program to clean up the memory.
    This means that latency or pauses is introduced to every program, and can occur at any point in the execution of the program, such as during the hot path of the program.
    This has led to garbage collected languages being unsuitable for use in systems programs such as OS's, as for example it would be undesirable to have seemingly random latency when interacting with the system.
    There has been research in modern garbage collection algorithms such as ZGC~\cite{ZGC} and Shenandoah~\cite{flood2016shenandoah} for Java which aim to reduce the number and duration of pauses where the garbage collector runs, increasing the responsiveness of the Java virtual machine.
    However, even with much reduced duration of pauses, it may still be unsuitable for some systems with strict latency requirements such as OS's or financial systems.

@@

    This design mirrors the modular design of LLVM where because the analysis is performed on the IR, it could be reused by the frontend of a different language and not necessarily Clang as we do.
    Secondly this also enables us to flexibly test different algorithms as the C++ specific parsing and refactoring code is not affected by the choice or power of the analysis.
    This design has been already used to create a prototype of the tool~\footnote{https://github.com/a-usov/ptr-tidy/tree/c44e67022ba5fa2b3a541f4e730230c1e378e46e}.
    The prototype shows that combining the AST and IR is viable and uses the already implemented LLVM escape analysis on variable declarations.
    It also shows how Clang library allows for easy refactoring since the AST stores all source text and line numbers, making text substitution easy.
    Substitution of AST nodes rather than source text substitution has not been explored, however this is as documentation states that generally the AST should remain immutable.

@@

    Show how you plan to organize your work, identifying intermediate deliverables and dates.

    \begin{landscape}
        \begin{figure}
            \begin{ganttchart}{1}{14}
                \gantttitle{Work Plan}{14} \ganttnewline
                \gantttitlelist{1,...,14}{1} \ganttnewline
                \ganttgroup{Complete Prototype}{1}{3} \ganttnewline
                \ganttbar{Finish Rewriting Component}{1}{1} \ganttnewline
                \ganttbar{Combine all components into final prototype}{2}{2} \ganttnewline
                \ganttbar{Evaluate final prototype and find edge-cases}{3}{3} \ganttnewline
                \ganttgroup{Benchmark and Data Acquisition}{2}{3}  \ganttnewline
                \ganttbar{Find, setup and test benchmark suite}{2}{3} \ganttnewline
                \ganttgroup{Analysis Algorithm Exploration}{4}{7} \ganttnewline
                \ganttbar{Implement proposed interprocedural analysis}{4}{5}\ganttnewline
                \ganttlink{elem3}{elem7}
                \ganttbar{Evaluate tool with interprocedural analysis}{6}{6} \ganttnewline
                \ganttbar{Implement cycle analysis and evaluate}{7}{7} \ganttnewline
                \ganttgroup{Result Consolidation}{8}{14} \ganttnewline
                \ganttbar{Research Paper Writeup}{8}{14} \ganttnewline
            \end{ganttchart}
            \caption{[A Gantt chart demonstrating the project schedule, broken down over the first
            14 weeks of the year from Monday 11th January until Friday 16th April.
            ]}
            \label{fig:gantt}
        \end{figure}
    \end{landscape}

>>>>

    \title{ptr-tidy: Automatic refactoring of raw pointers in C++}
    \author{Artem Usov\\ 2296905U}
    \date{18 December 2020}
    \maketitle
    \tableofcontents
    \newpage

@@

    Memory safety errors occur when languages are used that place the tasks of memory management with the programmer, such as C and C++.
    However simply not using these languages is not an option since their unmanaged nature makes them highly performant, and therefore the best option for a system such as a web browser.

@@

    Engineers at both Microsoft~\cite{Thomas2019} and Mozilla~\cite{Hostfelt2019} converge on Rust~\cite{Balasubramanian2017} as a possible solution.
    Rust is a systems language that offers the same performance as C and C++~\cite{Lin2016}, however its linear type system and memory ownership model also guarantee memory safety.
    The rewriting of a program in a new language is a colossal undertaking, especially in a language that has a reputation for being difficult to learn.
    We therefore propose an alternative partial solution to this problem by using a concept from Rust.

@@

    Generally pointers are used to give the program access to a resource that cannot be directly included in the program itself, such as a file or a thread.
    However Stroustrup claims that pointers to objects allocated on the free store are dangerous and a \emph{plain old pointer}, or raw pointer as we will refer to them, should not be used to represent ownership~\cite{Stroustrup2018}.

@@

    In \autoref{lst:manual-leak} we can see that when using a raw pointer, there are two cases in which the programmer will never free the allocated memory.
    Instead we can use \emph{smart pointers}~\cite{Dimov2003} and the concept of \emph{RAII (Resource Acquisition Is Initialisation)} to create resource handles which automatically eliminate resource leaks with no added overhead.
    Firstly there are \emph{unique pointers} which are pointers that have only a single unique owner at any time.
    We can therefore fix the cases in which we leak the memory using a unique pointer which can be seen in in \autoref{lst:first-example-unique}.
    In the comments, we can see all the locations where \emph{p} exits the scope of the function and where the compiler will automatically insert delete statements, now ensuring that the resource will never get leaked.
    The ownership of a unique pointer can be transferred between variables using \mintinline{c++}{std::move} which moves the ownership from one variable to another.

@@

    There also exists a \emph{shared pointer} for resources which may not have a single unique owner.
    Reference counting is used to ensure that after the last owner loses access to the resource, it will finally be freed.
    Shared pointers are suitable to be used in most situations, except in structures which have cyclic references \autoref{subsec:shared-pointer-cycle-analysis} and such reference counted pointers are used by default for all values in Python and Swift.

@@

    % TODO find refernce of how much old code is in production

    We therefore propose to create an automatic refactoring tool \emph{(ptr-tidy)}, that statically analyses C++ code and aims to refactor usage of raw pointers into smart pointers.
    It will involve a conservative analysis that first and foremost tries to identify raw pointers that only have a single unique owner at any point in the program.
    These will be refactored into the previously mentioned unique pointers.
    If this cannot be guaranteed, then a more conservative refactoring to a shared pointer will be applied.
    Given that the language standards are mostly backwards-compatible~\cite{Winters2020}, we can apply these automatic refactoring without changing the runtime behaviour compared to the original programs.

    We use the term refactoring as this is one most familiar to programmers to convey a change in the code that does not change the behaviour, but improves the quality or maintainability of the code.
    However, our work aligns more closely with the term \emph{source code rejuvenation}, as defined by \citet{Pirkelbauer2010}.
    This is because we exactly aim to replace outdated coding patterns with newer, higher-level abstractions, as well fitting the notion that this tool would be applied once to a codebase, rather than be a reoccurring task.
    The two terms can be used interchangeably in the context of this paper, however we will exclusively use the term refactoring.

@@

    \emph{"Can the translation of raw pointers into smart pointers be used to make existing C++ code more modern and extendable?"}

    \emph{"Can the tool reliably identify situations in which unique pointers should be used?"}

@@

    C++ as a language is massive, and the first capability our tool will need is to be able to parse C++ code.
    Doing this ourselves would be an undertaking that would likely span longer than the time span of the whole project, therefore it was decided early to use an existing parser.
    The most capable and mature C++ parsers will naturally be the parsers used by the most popular compilers.
    This leaves us therefore with a choice between GCC~\footnote{\url{https://gcc.gnu.org/}} and Clang~\footnote{\url{https://clang.llvm.org/}}.
    The literature~\cite{Duffy2014} is fairly clear on the fact that the design of GCC makes it unsuitable for integration with other projects.
    Clang on the other hand allows us access to the Clang abstract syntax tree (AST) and parser internals through standard design patterns such as consumers and visitors.

    Clang is a front-end to the LLVM framework~\cite{Lattner2004} which transforms C++ code into the LLVM intermediate representation (IR).
    The LLVM IR is a language-independent, static single assignment (SSA)~\cite{Rosen1988} representation of the program and will be the target on which we apply our analysis to identify refactoring opportunities.
    We believe this choice will lead to several benefits in terms of the analysis power and future usage of the project:
    \begin{itemize}
        \item First, the usage of SSA allows us to efficiently generate a definition-usage graph, therefore allowing an efficient analysis of all the usages of a pointer to identify its number of owners.
        \item Secondly, by separating the parsing and analysis, we make our analysis stage reusable by any other LLVM frontends as the analysis is source language agnostic.
    \end{itemize}

@@

    \begin{listing}
        \inputminted{c++}{code/code-to-ast.cpp}
        \inputminted[fontsize=\footnotesize]{text}{code/code-to-ast.txt}
        \centerline{\includegraphics{images/code-to-ast}}
        \caption{Example of converting C++ code to Clang AST. All diagrams show equivalent representation of a small C++ code snippet.}
        \label{lst:code-to-ast-example}
    \end{listing}

@@

    The escape analysis algorithm described by \citet{Choi1999} is used as the default escape analysis algorithm in the Oracle Java Virtual Machine implementation~\footnote{\url{https://docs.oracle.com/en/java/javase/15/vm/java-hotspot-virtual-machine-performance-enhancements.html}} and is widely cited so forms the basis of our investigation.
    Due to how similar C++ and Java are in terms of their Object-Oriented features, will analyse their algorithm as well as the implementations of escape analysis in the LLVM and Go compilers.

@@

    Go is another language that uses a escape analysis to allocate objects on the stack rather than the heap.
    Their escape analysis algorithms is not defined in the specification of the language, however we can look at the source~\footnote{\url{https://github.com/golang/go/blob/master/src/cmd/compile/internal/gc/escape.go}} to determine what algorithm they use.
    Fortunately the comments give a summary of their algorithm, which involves constructing a weighted graph where vertices represent variables, and edges represent assignments between variables.
    They walk the graph, looking for assignment paths that violate the invariants of a variable being allocated on the stack and mark those as requiring heap allocation
    We can see that this approach is highly similar to the use of connection graphs by \citet{Choi1999}, however there is lacking documentation on how they perform interprocedural analysis so their analysis is perhaps weaker.

    Finally, LLVM also uses escape analysis to remove any unnecessary heap allocations.
    The escape analysis algorithm is also not formally defined, however we can also look at the source code~\footnote{\url{https://github.com/llvm/llvm-project/blob/master/llvm/lib/Analysis/CaptureTracking.cpp}}.
    We see that the \mintinline{c++}{PointerMayBeCaptured} function uses the fact that the single static assignment (SSA) in the LLVM IR allows for efficient creation of definition-usage chains.
    For any defined value, such as a value initialising a variable, we can traverse all its usages, such as copying to another variable.
    To determine if an object escapes, we analyse all its usages, and all their usages and so on.
    For each usage, we then determine if it is an instruction which would cause the object to escape.
    This leads to a roughly similar analysis as the work by \citet{Choi1999} when considering the intraprocedural analysis alone, as we informally explore a graph of the nodes that represent objects and nodes that obtain references to objects by using the original value in a copy instruction.
    However, its assumption about different instructions are conservative, and it also has no interprocedural analysis, making it the weakest analysis so far.

@@

    Swift remains mostly a language for the Apple MacOS and iOS environments for application creation, however it has seen some usage for system programs by companies such as IBM~\footnote{\url{https://github.com/ibm-swift}}.
    Ada or its subset language SPARK are used extensively in certain industries to create verifiable programs for critical systems such as in avionics.
    It is perhaps most well-know for being used in the F-22 fighter aircraft~\cite{Malas1997}, however the use of Ada for those systems show its other weakness in productivity time as it was the cause for several delays in the project, and so it largely remains unused in areas which do not call for such meticulous safety.

    Rust has been the most popular language in attractive widespread adoption as a systems language.
    Its popularity could be quantified by considering that it is being considered for usage in writing Linux kernel modules~\footnote{\url{https://lore.kernel.org/lkml/CAKwvOdmuYc8rW_H4aQG4DsJzho=F+djd68fp7mzmBp3-wY--Uw@mail.gmail.com/T}}, which cannot be said for C++.
    Following the advice given by many that most systems programs should be written in Rust from now on, we see a potential future usage of our work in the creation of a C++ to Rust transpiler.

@@

    Garbage collected languages encompass some of the most popular programming languages such as Java and Python.
    They are considered to be simpler languages than C++ since they remove the need for manual memory management by the programmer.
    In C++, if memory is improperly freed, it gets leaked and can eventually cause out of memory errors.
    Garbage collectors instead run periodically and scan the memory of the program to find objects in memory which were allocated but are no longer used anymore, i.e leaked and then free the leaked objects.
    Languages such as Java~\cite{Tauro2012} use this fully by allocating almost every object on the heap and not freeing any object, instead relying on the garbage collector to clean up.
    They are therefore considered to be easier and more productive to use.

    Garbage collection does also exist for C++ such as the Boehm-Demers-Weiser conservative garbage collector~\cite{Boehm2002}.
    So why is garbage collection not used in every language?
    Garbage collection is not zero-cost as the garbage collection algorithm typically stops the execution of the running program to clean up the memory.
    This means that latency or pauses is introduced to every program, and can occur at any point in the execution of the program, such as during the hot path of the program.
    This has led to garbage collected languages being unsuitable for use in systems programs such as OS's, as for example it would be undesirable to have seemingly random latency when interacting with the system.
    There has been research in modern garbage collection algorithms such as ZGC~\cite{Liden2018} and Shenandoah~\cite{Flood2016} for Java which aim to reduce the number and duration of pauses where the garbage collector runs, increasing the responsiveness of the Java virtual machine.
    However, even with much reduced duration of pauses, it may still be unsuitable for some systems with strict latency requirements such as OS's or any other real time systems.

    \subsection{Existing Tools and Research}\label{subsec:existing-tools-and-research}

    As we have mentioned in \autoref{subsec:do-smart-pointer-solve-memory-management?}, our work closely aligns with the research done around \emph{source code rejuvenation}.
    We believe research in this area is going to increase, as languages move and evolve much quicker than can be supported in enterprise environments, which is supported by the fact that other relevant work has been completed on this subject.

    First and foremost, we have evidence that this is a problem that actually exists in industry through a paper by \citet{Wright2013} which describes a tool used in Google for running such rejuvenations across a large codebase and shows an example of running a simple rejuvenation of upgrading an API call.
    It only uses Clang for the parsing of code into an AST, and traverses the tree to perform simple refactorings.
    This allows for less complex refactorings, however it allows for a flexible design to implement different refactorings using the same tool, rather than the focus we have for a single specific refactoring goal.

    Secondly, there has been separate academic research into rejuvenation tools such as the work by \citet{Huck2016} and \citet{Kumar2012}.
    The former also uses the Clang and AST driven approach that we have described in \autoref{subsec:clang-and-llvm} and by \citet{Wright2013}, whilst the latter uses a representation called IPR~\cite{Dos2011}, which is a general and efficient data structure for representing C++ programs.
    We have not evaluated IPR as a representation to use instead of the Clang AST, but we can however comment that whilst we appreciate the aim of the project to create a efficient and compiler-independent representation, the greater amount of documentation that Clang has, as well as it being the choice of several other successful projects leads us to believe that we will not lose any tool efficiency in our use of it.

@@

    This design mirrors the modular design of LLVM where because the analysis is performed on the IR, it could be reused by the frontend of a different language and not necessarily Clang as we do.
    Secondly this also enables us to flexibly test different algorithms as the C++ specific parsing and refactoring code is not affected by the choice or power of the analysis.
    This design has been already used to create a prototype of the tool~\footnote{\url{https://github.com/a-usov/ptr-tidy/tree/c44e67022ba5fa2b3a541f4e730230c1e378e46e}}.
    The prototype shows that combining the AST and IR is viable and uses the already implemented LLVM escape analysis on variable declarations.
    It also shows how Clang library allows for easy refactoring since the AST stores all source text and line numbers, making text substitution easy.
    Substitution of AST nodes rather than source text substitution has not been explored, however this is as documentation states that generally the AST should remain immutable.

@@

    \subsection{Risk Analysis}\label{subsec:risk-analysis}

    We have performed a risk analysis for the project, our chosen implementation as well as the work plan we envision for the completion of this project.
    \begin{itemize}
        \item First, the frameworks we have chosen to use, namely LLVM and Clang are mature and enterprise-backed open-source frameworks, and therefore we see no risk in using them for current and any future development,
        \item Second, the work plan has been structured in such a way that given any unforeseen setbacks in the project, our existing prototype can be finished without the integration of more complex analyses to provide a minimum viable product which can be used in experimentation and evaluation.
        This ensures that we can answer our research questions for this project.
    \end{itemize}

    % TODO - finish this

    \subsection{Gantt Chart}\label{subsec:gantt-chart}

    \begin{landscape}
        \begin{figure}
            \begin{ganttchart}{1}{14}
                \gantttitle{Work Plan}{14} \ganttnewline
                \gantttitlelist{1,...,14}{1} \ganttnewline
                \ganttgroup{Complete Prototype}{1}{3} \ganttnewline
                \ganttbar{Finish Rewriting Component}{1}{1} \ganttnewline
                \ganttbar{Combine all components into final prototype}{2}{2} \ganttnewline
                \ganttbar{Evaluate final prototype and find edge-cases}{3}{3} \ganttnewline
                \ganttgroup{Benchmark and Data Acquisition}{2}{3}  \ganttnewline
                \ganttbar{Find, setup and test benchmark suite}{2}{3} \ganttnewline
                \ganttgroup{Analysis Algorithm Exploration}{4}{7} \ganttnewline
                \ganttbar{Implement proposed interprocedural analysis}{4}{5}\ganttnewline
                \ganttlink{elem3}{elem7}
                \ganttbar{Evaluate tool with interprocedural analysis}{6}{6} \ganttnewline
                \ganttbar{Implement cycle analysis and evaluate}{7}{7} \ganttnewline
                \ganttgroup{Result Consolidation}{8}{14} \ganttnewline
                \ganttbar{Research Paper Writeup}{8}{14} \ganttnewline
            \end{ganttchart}
            \caption{A Gantt chart demonstrating the project schedule, broken down over the first
            14 weeks of the year from Monday 11th January until Friday 16th April.
            }
            \label{fig:gantt}
        \end{figure}
    \end{landscape}
